Long- and Short-Read Metabarcoding Give Similar Results for Spatial Turnover Range in an Ectomycorrhizal Savanna Woodland

### 2020-02-14 writing/transect_paper.Rmd
Move methods for Illumina denoising to end of section, clarify that this was different from PacBio/Ion Torrent and why; mention problems with DNA extracts from Ang 2015.

# use the "light" plan cache instead of the default cache at .drake
cache <- drake_cache(here(".light"))
readd <- function(..., cache = drake_cache(here(".light"))) {
  force(cache)
  drake::readd(..., cache = cache)
}
loadd <- function(..., cache = drake_cache(here(".light"))) {
  force(cache)
  drake::loadd(..., cache = cache, envir = parent.frame())
}

@@

Forward and reverse Illumina reads were truncated after the first base with quality score <= 10, and then reads with more than 3 expected errors in either read were discarded.
Forward and reverse reads were denoised using `DADA2` version `r packageVersion("dada2")` [@callahan2016] using seperate error models and pseudo-pooling, and then forward and reverse reads were merged.

We attempted to denoise both long and short PacBio amplicons using `DADA2` according to the steps outlined in  the supplementary information in @callahan2019.
However, only `r readd(region_table)$ASVs[readd(region_table)$region=="long"]` amplicon sequence variants (ASVs) were obtained for the long amplicons, representing `r percent(readd(region_table)$reads[readd(region_table)$region == "long"]/as.integer(gsub(" ", "", readd(demuxcounts)["pb_500"])), format = "d")` of the trimmed reads.
We believe that this poor performance was due to a combination of long read length and low sequencing depth relative to community diversity.
The `DADA2` algorithm requires that the seed sequence of each ASV be represented by at least two error-free reads [@callahan2016].
If sequencing errors are uniformly distributed, then the probability that a given read will be error-free is $(1-\epsilon)^L$, where $\epsilon$ is the sequencing error rate and $L$ is the read length in base pairs.
Then the number of reads of a given sequence that would be required to obtain two error-free reads in expectation is $2/(1-\epsilon)^L$.
For the combination of long reads (median $L=`r readd(demuxlength)['pb_500']` \text{ bp}$ after trimming) and moderate error rate (mean $\epsilon= `r format(readd(demuxqual)['pb_500'], digits = 4)`$ based on ccs quality scores) for the long amplicon in this study, the expected number of reads required to achieve two error-free reads is `r format(as.integer(round(2/(1 - readd(demuxqual)['pb_500'])^readd(demuxlength)['pb_500'])), big.mark = " ")`.
Given the high diversity relative to sequencing depth in this study (`r readd(region_table)$ASVs[readd(region_table)$seq_run == "pb_483" & readd(region_table)$region == "short"]` ASVs based on PacBio short amplicons, `r readd(demuxcounts)["pb_500"]` trimmed long amplicon reads), this requirement could not have been met for the long amplicons except by the most abundant sequences.
In comparison, the equivalent requirement for the short amplicon ($L=`r readd(demuxlength)['pb_483']` \text{ bp}$, $\epsilon=`r readd(demuxqual)['pb_483']`$) is only `r round(2/(1 - readd(demuxqual)['pb_483'])^readd(demuxlength)['pb_483'], 1)` reads.
We therefore developed a new workflow to assemble ASVs from the long amplicons, as follows:

@@

@@

The number of sequencing reads and ASVs at each stage in the bioinformatics pipeline are shown in Table\ \@ref(tab:bioinfo).
Long (short) amplicon sequencing with PacBio yielded `r readd(rawcounts)["pb_500"]` (`r readd(rawcounts)["pb_483"]`) circular consensus reads, of which `r readd(demuxcounts)["pb_500"]` (`r readd(demuxcounts)["pb_483"]`) were successfully trimmed and demultiplexed.
Illumina sequencing yielded `r readd(rawcounts)["SH-2257"]` read pairs, of which `r readd(demuxcounts)["SH-2257"]` were successfully trimmed and demultiplexed, and `r readd(filtercounts_full)["SH-2257"]` passed quality filtering.
Ion Torrent sequencing yielded `r readd(demuxcounts)["is_057"]` demultiplexed reads.
`r readd(filtercounts_full)['pb_500']` (`r percent(readd(filtercounts_full)['pb_500']/readd(demuxcounts)['pb_500'], digits = 1, format = "f")`) of long amplicon PacBio reads passed length and quality filtering stages, compared with `r readd(filtercounts_full)['pb_483']` (`r percent(readd(filtercounts_full)['pb_483']/readd(demuxcounts)['pb_483'], digits = 1, format = "f")`) of short amplicon PacBio reads, `r readd(filtercounts_full)['SH-2257']` (`r percent(readd(filtercounts_full)['SH-2257']/readd(demuxcounts)['SH-2257'], digits = 1, format = "f")`) of Illumina read pairs, and `r readd(filtercounts_full)['is_057']` (`r percent(readd(filtercounts_full)['is_057']/readd(demuxcounts)['is_057'], digits = 1, format = "f")`) of Ion Torrent reads.

@@

```{r bioinfo}
#TODO add filtered reads (currently inflated numbers for non-Illumina)
#TODO asterisks for mapped ASVs
# loadd(bioinf_table)
options(knitr.kable.NA = '–')

readd(bioinf_table) %>%
  set_names(gsub(".*_", "", names(.))) %>%
  kable(
    booktabs = TRUE,
    caption = "(ref:readcounts)",
    align = "rrrrrr",
    format.args = list(big.mark = " "),
    linesep = c(rep("", 3), rep("\\addlinespace", 2), rep("", 4)),
    escape = FALSE
    ) %>%
  kableExtra::add_header_above(
    header = c(
      " " = 1,
      readd(bioinf_header)$amplicon[c(TRUE, FALSE)] %>% set_names(rep(2, length(.)), .)
    )
  ) %>%
  kableExtra::add_header_above(
    header = c(
      " " = 1,
      rle(readd(bioinf_header)$tech) %$% set_names(lengths, values)
    )
  )
```

@@

The 5.8S rDNA was located in `r readd(regioncounts)['pb_500']` long amplicon PacBio reads (`r percent(readd(regioncounts)['pb_500']/readd(demuxcounts)['pb_500'], digits = 1, format = "f")` of demultiplexed total), `r readd(regioncounts)['pb_483']` short amplicon PacBio reads (`r percent(readd(regioncounts)['pb_483']/readd(demuxcounts)['pb_483'], digits = 1, format = "f")`), and `r readd(regioncounts)['is_057']` Ion Torrent reads (`r percent(readd(regioncounts)['is_057']/readd(demuxcounts)['is_057'], digits = 1, format = "f")`).
Length and quality filtering resulted in `r readd(filtercounts_ITS2)['pb_500']` (`r percent(readd(filtercounts_ITS2)['pb_500']/readd(demuxcounts)['pb_500'], digits = 1, format = "f")` of demultiplexed total) ITS2 sequences from the long amplicon PacBio reads, `r readd(filtercounts_ITS2)['pb_483']` (`r percent(readd(filtercounts_ITS2)['pb_483']/readd(demuxcounts)['pb_483'], digits = 1, format = "f")`) ITS2 sequences from the short amplicon PacBio reads, and `r readd(filtercounts_ITS2)['is_057']` (`r percent(readd(filtercounts_ITS2)['is_057']/readd(demuxcounts)['is_057'], digits = 1, format = "f")`) ITS2 sequences from the Ion torrent reads.
The length distribution of different regions extracted from the long amplicon are shown in Supplementary Figure\ \@ref(fig:region-lengths).

@@

```{r venn, fig.cap="(ref:venn)", fig.show="hold", out.width = "90%"}
venn_A <- venndata(readd(asv_table), ASVs, cols = c("pb_483", "SH-2257", "is_057")) %>%
  vennplot_data(ASVs) %>% {
    plot(
      select(., set, ASVs) %>% deframe() %>% replace_na(0) %>% eulerr::venn(),
      legend = FALSE,
      labels = FALSE,
      quantities = .$label,
      main = "a",
      fills = c("grey85", "lightblue", "lightcoral"))
  }

venn_B <- venndata(readd(otu_table), OTUs, cols = c("pb_483", "SH-2257", "is_057")) %>%
  vennplot_data(OTUs) %>% {
    plot(
      select(., set, OTUs) %>% deframe() %>% replace_na(0) %>% eulerr::venn(),
      legend = FALSE,
      labels = FALSE,
      quantities = .$label,
      main = "b",
      fills = c("grey85", "lightblue", "lightcoral")
    )
  }
ggpubr::ggarrange(venn_A, venn_B, ncol = 2)

@@

venn_C <- readd(asv_table) %>%
  mutate_if(is.numeric, ~ . / sum(.)) %>%
  mutate(long = pb_500,
         short = rowMeans(.[,c("pb_483", "SH-2257", "is_057")])) %>%
  select(seq, long, short) %>%
  venndata(ASVs, cols = c("long", "short")) %>%
  vennplot_data(ASVs) %>% {
    plot(
      select(., set, ASVs) %>% deframe() %>% replace_na(0) %>% eulerr::venn(),
      legend = FALSE,
      labels = FALSE,
      quantities = .$label,
      main = "c",
      fills = c("white", "slateblue1"))
  }

venn_D <- readd(otu_table) %>%
  mutate_if(is.numeric, ~ . / sum(.)) %>%
  mutate(long = pb_500,
         short = rowMeans(.[,c("pb_483", "SH-2257", "is_057")])) %>%
  select(seq, long, short) %>%
  venndata(OTUs, cols = c("long", "short")) %>%
  vennplot_data(OTUs) %>% {
    plot(
      select(., set, OTUs) %>% deframe() %>% replace_na(0) %>% eulerr::venn(),
      legend = FALSE,
      labels = FALSE,
      quantities = .$label,
      main = "d",
      fills = c("white", "slateblue1")
    )
  }
# ggpubr::ggarrange(venn_C, venn_D, ncol = 2)
# grid::grid.legend(
#   labels =  c("Long amplicon", "Short amplicon"),
#   nrow = 1,
#   pch = 21,
#   gp = grid::gpar(fill = c("white", "slateblue1")),
#   vp = grid::viewport(y = unit(0.2, "npc"))
# )
```

@@

```{r pb-vs-it, fig.cap="(ref:readcap)", fig.height=8}
ggpubr::ggarrange(
  read_comparison(
    readd(multi_table),
    readd(comparisons),
    type = "ASV"
  ),
  read_comparison(
    readd(multi_table),
    readd(comparisons),
    type = "OTU"
  ),
  ncol = 1,
  labels = "auto",
  label.x = 0.9
)
```

The number of ASVs and OTUs shared between the different sequencing strategies are shown in Supplementary Tables \@ref(tab:venn-table-asv) and \@ref(tab:venn-table-otu), respectively, and graphically in Figure\ \@ref(fig:venn).
ASVs shared between all datasets represented `r text_list(percent(readd(venn_ASV)["1111", c("frac_pb_500", "frac_pb_483", "frac_SH-2257", "frac_is_057")]))` of the reads for the long and short PacBio datasets, Illumina dataset, and Ion Torrent dataset, respectively.
These fractions increased to `r text_list(percent(readd(venn_OTU)["1111", c("frac_pb_500", "frac_pb_483", "frac_SH-2257", "frac_is_057")]))` when differences at the intra-species scale were removed by clustering the ASVs into 97% OTUs.
In particular, `r text_list(percent(colSums(mutate_all(readd(venn_OTU)[c("0111", "1111"), c("frac_pb_483", "frac_SH-2257", "frac_is_057")], as.numeric), na.rm = TRUE)))` of reads in the PacBio, Illumina, and Ion Torrent short-amplicon datasets belonged to OTUs shared between all three datasets.
In contrast, `r percent(readd(venn_ASV)["1000", "frac_pb_500"])` of reads in the long PacBio dataset belonged to ASVs which were unique to that dataset, and the fraction only reduced to `r percent(readd(venn_OTU)["1000", "frac_pb_500"])` after OTU clustering.

@@

Almost all of the short amplicons from all three technologies were between 240 and 375 bp long (Supplementary Figure\ \@ref(fig:full-length)a).
Although the length profile of the three requencing runs were very similar, Illumina MiSeq had the largest fraction of reads near the top of the range, followed by Ion Torrent Ion S5 and PacBio RS II (Supplementary Figure\ \@ref(fig:full-length)b).
In contrast, the length of the long amplicon reads varied widely, from `r filter(readd(reads_table), region == "long") %$% range(length) %>% paste(collapse = " to ")` bp, with a median of `r filter(readd(reads_table), region == "long") %$% reldist::wtd.quantile(length, weight = reads)` bp.

@@

```{r tax-chart, fig.cap = "(ref:tax-chart)"}
readd(tax_chart_plot)
```

@@

```{r taxa, fig.cap = "(ref:taxa)", out.width="100%", fig.width = 8}
fungi_read <- readd(taxdata) %>%
  taxa::filter_taxa(
    taxon_names == "Fungi",
    reassign_obs = FALSE,
    reassign_taxa = FALSE
    ) %>%
  extract2("data") %>%
  extract2("tax_read") %>%
  select(ends_with("PHYLOTAX")) %>%
  set_names(names(.) %>% str_extract("(PacBio|Illumina|Ion.Torrent)_(Long|Short)")) %>%
  unlist()

fungi_asv <- readd(taxdata) %>%
  taxa::filter_taxa(
    taxon_names == "Fungi",
    reassign_obs = FALSE,
    reassign_taxa = FALSE
    ) %>%
  extract2("data") %>%
  extract2("tax_asv") %>%
  select(ends_with("PHYLOTAX")) %>%
  set_names(names(.) %>% str_extract("(PacBio|Illumina|Ion.Torrent)_(Long|Short)")) %>%
  unlist()

ecm_counts <- readd(taxon_reads) %>%
  filter(Algorithm == "PHYLOTAX") %>%
  group_by(seq_run) %>%
  mutate(ASVs = 1/n()) %>%
  filter(ECM %in% c("Possible ECM", "Probable ECM", "Highly Probable ECM")) %>%
  summarize_at(c("ASVs", "reads"), sum)

@@

theme <- theme_update(panel.border = element_blank())
readd(heattree)
theme_set(theme)
```

@@

```{r heattree-db-compare, fig.cap = "(ref:heattree-db-compare)", include = FALSE}
taxdata2 <- readd(taxdata)
# taxdata2$data$asv_table <- metacoder::calc_obs_props(taxdata2, "tax_data", cols = asv_cols)
# taxdata2$data$read_table <- metacoder::calc_obs_props(taxdata2, "tax_data", cols = read_cols)
# taxdata2$data$tax_asv <- metacoder::calc_taxon_abund(taxdata2, "asv_table", cols = asv_cols)
# taxdata2$data$tax_read <- metacoder::calc_taxon_abund(taxdata2, "read_table", cols = read_cols)
taxdata2$data$diff_read <- metacoder::compare_groups(
  taxdata2,
  data = "tax_read",
  cols = names(taxdata2$data$tax_read) %>%
    keep(str_detect, "reads_PacBio_Long_(Unite|RDP)"),
  groups = names(taxdata2$data$tax_read) %>%
    keep(str_detect, "reads_PacBio_Long_(Unite|RDP)") %>%
    str_replace("reads_.+_.+_(.+)_.+", "\\1"),
  func = function(abund1, abund2) {
    list(read_ratio = log10(mean(abund1) / mean(abund2)))
  }
  )
taxdata2$data$diff_asv <- metacoder::compare_groups(
  taxdata2,
  data = "tax_asv",
  cols = names(taxdata2$data$tax_asv) %>% keep(startsWith, "ASVs_"),
  groups = names(taxdata2$data$tax_asv) %>%
    keep(startsWith, "ASVs_") %>%
    str_replace("ASVs_.+_(.+)_.+_.+", "\\1"),
  func = function(abund1, abund2) {
    list(asv_ratio = log10(mean(abund1) / mean(abund2)))
  }
)

@@

*Agaricus bisporus*, the positive control, was represented by a single ASV in the positive control samples for both long- and short-amplicon PacBio datasets, and in the Ion Torrent dataset.
Additional ASVs belonging to the genus *Agaricus* were also detected in the same wells as the positive control in the Ion Torrent dataset, but these were sufficiently different that they probably represent additional *Agaricus* species from the field samples, and were found in the same wells due to incomplete demultiplexing.
*A. bisporus* was represented by two ASVs in the Illumina dataset, which differed at one base pair.
The abundance of the second ASV was `r readd(agaricus_reads) %>% as.data.frame() %>% {.[["f2ab49e8"]]/.[["39d80c19"]]} %>% keep(~. > 0) %>% percent(format = "f", digits = 2) %>% paste(collapse = " and ")` that of the primary ASV in the two Illumina positive controls.
The consistency of this ratio across PCR replicates suggests that it represents true inter-copy variation within the specimen, rather than sequencing or PCR error.

The primary *A. bisporus* ASV was the most abundant ASV in the long-amplicon PacBio dataset and the Ion Torrent dataset, and the fifth most abundant ASV in the short-amplicon PacBio dataset, and the third most abundant ASV in the Illumina dataset.
This ASV was also found in `r readd(pos_control_data) %>% filter(!is.na(pc_frac)) %>% group_by(seq_run) %>% summarize(well_frac = sum(pc_frac > 0) / n()) %>% deframe() %>% magrittr::extract(c("pb_500", "pb_483", "SH-2257", "is_057")) %>% percent(format = "f", digits = 1) %>% text_list()` of the non-control samples, and represented `r readd(pos_control_data) %>% filter(complete.cases(.)) %>% group_by(seq_run) %>% summarize(read_frac = sum(pc_reads) / sum(all_reads)) %>% deframe() %>% magrittr::extract(c("pb_500", "pb_483", "SH-2257", "is_057")) %>% percent(format = "f", digits = 2) %>% text_list()` of non-control reads, in the PacBio long, PacBio short, Illumina, and Ion Torrent datasets, respectively.

@@

```{r phylotax, fig.height = 2, fig.width = 3, fig.cap = "(ref:phylotax)"}
readd(phylotax_fig)
```

@@

```{r region-lengths, fig.cap = "(ref:region-lengths)", fig.height = 6}
readd(region_length_fig)
```

>>>>

# use the "light" plan cache instead of the default cache at .drake
cache <- drake_cache(here(".light"))
rdd <- function(..., cache = drake_cache(here(".light"))) {
  force(cache)
  drake::readd(..., cache = cache)
}
ldd <- function(..., cache = drake_cache(here(".light"))) {
  force(cache)
  drake::loadd(..., cache = cache, envir = parent.frame())
}

@@

We attempted to denoise both long and short PacBio amplicons using `DADA2` according to the steps outlined in  the supplementary information in @callahan2019.
However, only `r rdd(region_table)$ASVs[rdd(region_table)$region=="long"]` amplicon sequence variants (ASVs) were obtained for the long amplicons, representing `r percent(rdd(region_table)$reads[rdd(region_table)$region == "long"]/as.integer(gsub(" ", "", rdd(demuxcounts)["pb_500"])), format = "d")` of the trimmed reads.
We believe that this poor performance was due to a combination of long read length and low sequencing depth relative to community diversity.
The `DADA2` algorithm requires that the seed sequence of each ASV be represented by at least two error-free reads [@callahan2016].
If sequencing errors are uniformly distributed, then the probability that a given read will be error-free is $(1-\epsilon)^L$, where $\epsilon$ is the sequencing error rate and $L$ is the read length in base pairs.
Then the number of reads of a given sequence that would be required to obtain two error-free reads in expectation is $2/(1-\epsilon)^L$.
For the combination of long reads (median $L=`r rdd(demuxlength)['pb_500']` \text{ bp}$ after trimming) and moderate error rate (mean $\epsilon= `r format(rdd(demuxqual)['pb_500'], digits = 4)`$ based on ccs quality scores) for the long amplicon in this study, the expected number of reads required to achieve two error-free reads is `r format(as.integer(round(2/(1 - rdd(demuxqual)['pb_500'])^rdd(demuxlength)['pb_500'])), big.mark = " ")`.
Given the high diversity relative to sequencing depth in this study (`r rdd(region_table)$ASVs[rdd(region_table)$seq_run == "pb_483" & rdd(region_table)$region == "short"]` ASVs based on PacBio short amplicons, `r rdd(demuxcounts)["pb_500"]` trimmed long amplicon reads), this requirement could not have been met for the long amplicons except by the most abundant sequences.
In comparison, the equivalent requirement for the short amplicon ($L=`r rdd(demuxlength)['pb_483']` \text{ bp}$, $\epsilon=`r rdd(demuxqual)['pb_483']`$) is only `r round(2/(1 - rdd(demuxqual)['pb_483'])^rdd(demuxlength)['pb_483'], 1)` reads.
We therefore developed a new workflow to assemble ASVs from the long amplicons, as follows:

@@

Because the Illumina dataset consisted of paired-end reads, regions were not extracted prior to denoising.
ASVs were instead generated according to a standard workflow for `DADA2`.
Demultiplexed reads were truncated after the first base with quality score <= 10, and then reads with more than 3 expected errors in either read were discarded.
Forward and reverse reads were denoised using `DADA2` version `r packageVersion("dada2")` [@callahan2016] using seperate error models and pseudo-pooling, and then forward and reverse reads were merged.
The ITS2 region was extracted from the ASVs using `LSUx` for comparison to the other technologies.

@@

DNA concentrations after extraction and PCR, as well as sequencing reads for PacBio and Illumina, are shown per sample in Supplementary Figure \@ref(fig:sample_depth).
Samples from Ang in 2015 yielded low quantities of DNA, poor PCR performance, and ultimately very few sequencing reads, especially in the long amplicon library, where only one sample produced more than 100 reads.
Consequently Ang samples were excluded from spatial analysis, although they were retained for denoising, phylogenetic reconstruction, and taxonomic assignment.

The number of sequencing reads and ASVs at each stage in the bioinformatics pipeline are shown in Table\ \@ref(tab:bioinfo).
Long (short) amplicon sequencing with PacBio yielded `r rdd(rawcounts)["pb_500"]` (`r rdd(rawcounts)["pb_483"]`) circular consensus reads, of which `r rdd(demuxcounts)["pb_500"]` (`r rdd(demuxcounts)["pb_483"]`) were successfully trimmed and demultiplexed.
Illumina sequencing yielded `r rdd(rawcounts)["SH-2257"]` read pairs, of which `r rdd(demuxcounts)["SH-2257"]` were successfully trimmed and demultiplexed, and `r rdd(filtercounts_full)["SH-2257"]` passed quality filtering.
Ion Torrent sequencing yielded `r rdd(demuxcounts)["is_057"]` demultiplexed reads.
`r rdd(filtercounts_full)['pb_500']` (`r percent(rdd(filtercounts_full)['pb_500']/rdd(demuxcounts)['pb_500'], digits = 1, format = "f")`) of long amplicon PacBio reads passed length and quality filtering stages, compared with `r rdd(filtercounts_full)['pb_483']` (`r percent(rdd(filtercounts_full)['pb_483']/rdd(demuxcounts)['pb_483'], digits = 1, format = "f")`) of short amplicon PacBio reads, `r rdd(filtercounts_full)['SH-2257']` (`r percent(rdd(filtercounts_full)['SH-2257']/rdd(demuxcounts)['SH-2257'], digits = 1, format = "f")`) of Illumina read pairs, and `r rdd(filtercounts_full)['is_057']` (`r percent(rdd(filtercounts_full)['is_057']/rdd(demuxcounts)['is_057'], digits = 1, format = "f")`) of Ion Torrent reads.

@@

```{r bioinfo}
#TODO add filtered reads (currently inflated numbers for non-Illumina)
#TODO asterisks for mapped ASVs
options(knitr.kable.NA = '–')

rdd(bioinf_table) %>%
  set_names(gsub(".*_", "", names(.))) %>%
  kable(
    booktabs = TRUE,
    caption = "(ref:readcounts)",
    align = "rrrrrr",
    format.args = list(big.mark = " "),
    linesep = c(rep("", 3), rep("\\addlinespace", 2), rep("", 4)),
    escape = FALSE
    ) %>%
  kableExtra::add_header_above(
    header = c(
      " " = 1,
      rdd(bioinf_header)$amplicon[c(TRUE, FALSE)] %>% set_names(rep(2, length(.)), .)
    )
  ) %>%
  kableExtra::add_header_above(
    header = c(
      " " = 1,
      rle(rdd(bioinf_header)$tech) %$% set_names(lengths, values)
    )
  )
```

@@

The 5.8S rDNA was located in `r rdd(regioncounts)['pb_500']` long amplicon PacBio reads (`r percent(rdd(regioncounts)['pb_500']/rdd(demuxcounts)['pb_500'], digits = 1, format = "f")` of demultiplexed total), `r rdd(regioncounts)['pb_483']` short amplicon PacBio reads (`r percent(rdd(regioncounts)['pb_483']/rdd(demuxcounts)['pb_483'], digits = 1, format = "f")`), and `r rdd(regioncounts)['is_057']` Ion Torrent reads (`r percent(rdd(regioncounts)['is_057']/rdd(demuxcounts)['is_057'], digits = 1, format = "f")`).
Length and quality filtering resulted in `r rdd(filtercounts_ITS2)['pb_500']` (`r percent(rdd(filtercounts_ITS2)['pb_500']/rdd(demuxcounts)['pb_500'], digits = 1, format = "f")` of demultiplexed total) ITS2 sequences from the long amplicon PacBio reads, `r rdd(filtercounts_ITS2)['pb_483']` (`r percent(rdd(filtercounts_ITS2)['pb_483']/rdd(demuxcounts)['pb_483'], digits = 1, format = "f")`) ITS2 sequences from the short amplicon PacBio reads, and `r rdd(filtercounts_ITS2)['is_057']` (`r percent(rdd(filtercounts_ITS2)['is_057']/rdd(demuxcounts)['is_057'], digits = 1, format = "f")`) ITS2 sequences from the Ion torrent reads.
The length distribution of different regions extracted from the long amplicon are shown in Supplementary Figure\ \@ref(fig:region-lengths).

@@

```{r venn, fig.cap="(ref:venn)", fig.show="hold", out.width = "90%"}
venn_A <- venndata(rdd(asv_table), ASVs, cols = c("pb_483", "SH-2257", "is_057")) %>%
  vennplot_data(ASVs) %>% {
    plot(
      select(., set, ASVs) %>% deframe() %>% replace_na(0) %>% eulerr::venn(),
      legend = FALSE,
      labels = FALSE,
      quantities = .$label,
      main = "a",
      fills = c("grey85", "lightblue", "lightcoral"))
  }

venn_B <- venndata(rdd(otu_table), OTUs, cols = c("pb_483", "SH-2257", "is_057")) %>%
  vennplot_data(OTUs) %>% {
    plot(
      select(., set, OTUs) %>% deframe() %>% replace_na(0) %>% eulerr::venn(),
      legend = FALSE,
      labels = FALSE,
      quantities = .$label,
      main = "b",
      fills = c("grey85", "lightblue", "lightcoral")
    )
  }
ggpubr::ggarrange(venn_A, venn_B, ncol = 2)

@@

venn_C <- rdd(asv_table) %>%
  mutate_if(is.numeric, ~ . / sum(.)) %>%
  mutate(long = pb_500,
         short = rowMeans(.[,c("pb_483", "SH-2257", "is_057")])) %>%
  select(seq, long, short) %>%
  venndata(ASVs, cols = c("long", "short")) %>%
  vennplot_data(ASVs) %>% {
    plot(
      select(., set, ASVs) %>% deframe() %>% replace_na(0) %>% eulerr::venn(),
      legend = FALSE,
      labels = FALSE,
      quantities = .$label,
      main = "c",
      fills = c("white", "slateblue1"))
  }

venn_D <- rdd(otu_table) %>%
  mutate_if(is.numeric, ~ . / sum(.)) %>%
  mutate(long = pb_500,
         short = rowMeans(.[,c("pb_483", "SH-2257", "is_057")])) %>%
  select(seq, long, short) %>%
  venndata(OTUs, cols = c("long", "short")) %>%
  vennplot_data(OTUs) %>% {
    plot(
      select(., set, OTUs) %>% deframe() %>% replace_na(0) %>% eulerr::venn(),
      legend = FALSE,
      labels = FALSE,
      quantities = .$label,
      main = "d",
      fills = c("white", "slateblue1")
    )
  }
# ggpubr::ggarrange(venn_C, venn_D, ncol = 2)
# grid::grid.legend(
#   labels =  c("Long amplicon", "Short amplicon"),
#   nrow = 1,
#   pch = 21,
#   gp = grid::gpar(fill = c("white", "slateblue1")),
#   vp = grid::viewport(y = unit(0.2, "npc"))
# )
```

@@

```{r pb-vs-it, fig.cap="(ref:readcap)", fig.height=8}
ggpubr::ggarrange(
  read_comparison(
    rdd(multi_table),
    rdd(comparisons),
    type = "ASV"
  ),
  read_comparison(
    rdd(multi_table),
    rdd(comparisons),
    type = "OTU"
  ),
  ncol = 1,
  labels = "auto",
  label.x = 0.9
)
```

The number of ASVs and OTUs shared between the different sequencing strategies are shown in Supplementary Tables \@ref(tab:venn-table-asv) and \@ref(tab:venn-table-otu), respectively, and graphically in Figure\ \@ref(fig:venn).
ASVs shared between all datasets represented `r text_list(percent(rdd(venn_ASV)["1111", c("frac_pb_500", "frac_pb_483", "frac_SH-2257", "frac_is_057")]))` of the reads for the long and short PacBio datasets, Illumina dataset, and Ion Torrent dataset, respectively.
These fractions increased to `r text_list(percent(rdd(venn_OTU)["1111", c("frac_pb_500", "frac_pb_483", "frac_SH-2257", "frac_is_057")]))` when differences at the intra-species scale were removed by clustering the ASVs into 97% OTUs.
In particular, `r text_list(percent(colSums(mutate_all(rdd(venn_OTU)[c("0111", "1111"), c("frac_pb_483", "frac_SH-2257", "frac_is_057")], as.numeric), na.rm = TRUE)))` of reads in the PacBio, Illumina, and Ion Torrent short-amplicon datasets belonged to OTUs shared between all three datasets.
In contrast, `r percent(rdd(venn_ASV)["1000", "frac_pb_500"])` of reads in the long PacBio dataset belonged to ASVs which were unique to that dataset, and the fraction only reduced to `r percent(rdd(venn_OTU)["1000", "frac_pb_500"])` after OTU clustering.

@@

Almost all of the short amplicons from all three technologies were between 240 and 375 bp long (Supplementary Figure\ \@ref(fig:full-length)a).
Although the length profile of the three requencing runs were very similar, Illumina MiSeq had the largest fraction of reads near the top of the range, followed by Ion Torrent Ion S5 and PacBio RS II (Supplementary Figure\ \@ref(fig:full-length)b).
In contrast, the length of the long amplicon reads varied widely, from `r filter(rdd(reads_table), region == "long") %$% range(length) %>% paste(collapse = " to ")` bp, with a median of `r filter(rdd(reads_table), region == "long") %$% reldist::wtd.quantile(length, weight = reads)` bp.

@@

```{r tax-chart, fig.cap = "(ref:tax-chart)"}
rdd(tax_chart_plot)
```

@@

```{r taxa, fig.cap = "(ref:taxa)", out.width="100%", fig.width = 8}
fungi_read <- rdd(taxdata) %>%
  taxa::filter_taxa(
    taxon_names == "Fungi",
    reassign_obs = FALSE,
    reassign_taxa = FALSE
    ) %>%
  extract2("data") %>%
  extract2("tax_read") %>%
  select(ends_with("PHYLOTAX")) %>%
  set_names(names(.) %>% str_extract("(PacBio|Illumina|Ion.Torrent)_(Long|Short)")) %>%
  unlist()

fungi_asv <- rdd(taxdata) %>%
  taxa::filter_taxa(
    taxon_names == "Fungi",
    reassign_obs = FALSE,
    reassign_taxa = FALSE
    ) %>%
  extract2("data") %>%
  extract2("tax_asv") %>%
  select(ends_with("PHYLOTAX")) %>%
  set_names(names(.) %>% str_extract("(PacBio|Illumina|Ion.Torrent)_(Long|Short)")) %>%
  unlist()

ecm_counts <- rdd(taxon_reads) %>%
  filter(Algorithm == "PHYLOTAX") %>%
  group_by(seq_run) %>%
  mutate(ASVs = 1/n()) %>%
  filter(ECM %in% c("Possible ECM", "Probable ECM", "Highly Probable ECM")) %>%
  summarize_at(c("ASVs", "reads"), sum)

@@

theme <- theme_update(panel.border = element_blank())
rdd(heattree)
theme_set(theme)
```

@@

```{r heattree-db-compare, fig.cap = "(ref:heattree-db-compare)", include = FALSE}
taxdata2 <- rdd(taxdata)
# taxdata2$data$asv_table <- metacoder::calc_obs_props(taxdata2, "tax_data", cols = asv_cols)
# taxdata2$data$read_table <- metacoder::calc_obs_props(taxdata2, "tax_data", cols = read_cols)
# taxdata2$data$tax_asv <- metacoder::calc_taxon_abund(taxdata2, "asv_table", cols = asv_cols)
# taxdata2$data$tax_read <- metacoder::calc_taxon_abund(taxdata2, "read_table", cols = read_cols)
taxdata2$data$diff_read <- metacoder::compare_groups(
  taxdata2,
  data = "tax_read",
  cols = names(taxdata2$data$tax_read) %>%
    keep(str_detect, "reads_PacBio_Long_(Unite|RDP)"),
  groups = names(taxdata2$data$tax_read) %>%
    keep(str_detect, "reads_PacBio_Long_(Unite|RDP)") %>%
    str_replace("reads_.+_.+_(.+)_.+", "\\1"),
  func = function(abund1, abund2) {
    list(read_ratio = log10(mean(abund1) / mean(abund2)))
  }
  )
taxdata2$data$diff_asv <- metacoder::compare_groups(
  taxdata2,
  data = "tax_asv",
  cols = names(taxdata2$data$tax_asv) %>% keep(startsWith, "ASVs_"),
  groups = names(taxdata2$data$tax_asv) %>%
    keep(startsWith, "ASVs_") %>%
    str_replace("ASVs_.+_(.+)_.+_.+", "\\1"),
  func = function(abund1, abund2) {
    list(asv_ratio = log10(mean(abund1) / mean(abund2)))
  }
)

@@

*Agaricus bisporus*, the positive control, was represented by a single ASV in the positive control samples for both long- and short-amplicon PacBio datasets, and in the Ion Torrent dataset.
Additional ASVs belonging to the genus *Agaricus* were also detected in the same wells as the positive control in the Ion Torrent dataset, but these were sufficiently different that they probably represent additional *Agaricus* species from the field samples, and were found in the same wells due to incomplete demultiplexing.
*A. bisporus* was represented by two ASVs in the Illumina dataset, which differed at one base pair.
The abundance of the second ASV was `r rdd(agaricus_reads) %>% as.data.frame() %>% {.[["f2ab49e8"]]/.[["39d80c19"]]} %>% keep(~. > 0) %>% percent(format = "f", digits = 2) %>% paste(collapse = " and ")` that of the primary ASV in the two Illumina positive controls.
The consistency of this ratio across PCR replicates suggests that it represents true inter-copy variation within the specimen, rather than sequencing or PCR error.

The primary *A. bisporus* ASV was the most abundant ASV in the long-amplicon PacBio dataset and the Ion Torrent dataset, and the fifth most abundant ASV in the short-amplicon PacBio dataset, and the third most abundant ASV in the Illumina dataset.
This ASV was also found in `r rdd(pos_control_data) %>% filter(!is.na(pc_frac)) %>% group_by(seq_run) %>% summarize(well_frac = sum(pc_frac > 0) / n()) %>% deframe() %>% magrittr::extract(c("pb_500", "pb_483", "SH-2257", "is_057")) %>% percent(format = "f", digits = 1) %>% text_list()` of the non-control samples, and represented `r rdd(pos_control_data) %>% filter(complete.cases(.)) %>% group_by(seq_run) %>% summarize(read_frac = sum(pc_reads) / sum(all_reads)) %>% deframe() %>% magrittr::extract(c("pb_500", "pb_483", "SH-2257", "is_057")) %>% percent(format = "f", digits = 2) %>% text_list()` of non-control reads, in the PacBio long, PacBio short, Illumina, and Ion Torrent datasets, respectively.

@@

```{r phylotax, fig.height = 2, fig.width = 3, fig.cap = "(ref:phylotax)"}
rdd(phylotax_fig)
```

@@

```{r region-lengths, fig.cap = "(ref:region-lengths)", fig.height = 6}
rdd(region_length_fig)
```
